---
title: "Alpha diversity"
author: "Amina Hanaan"
date: "2026-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries 

```{r}
library(phyloseq)
library(ggpubr)
library(tidyverse)
library(readxl)

# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install("microbiome")

library(microbiome)
library(vegan)
# if (!requireNamespace("devtools", quietly = TRUE))
#   install.packages("devtools")

# devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")

library(pairwiseAdonis)
library(dplyr)
library(broom)
library(readr)
library(viridis)

```

#Loading data and metadata



```{r pressure, echo=FALSE}
set.seed(124)
prism_cb_safe <- c(
  # "#000000", 
  "#FB0165", 
  "#3297A7", 
  "#AB66FD", 
  "#41057E", 
  "#0072B2", 
  "#D55E00", 
  "#CC79A7"  
)

path = "."
setwd(path)
set.seed(124)

#====================================
#Loading data and metadata
#====================================

raw_data <- read.csv("Abundance_table.csv") %>%
  mutate(Kingdom = str_replace(Kingdom,                  
                               "Unclassified", "Others"))

#Metadata
metadata <- data.frame(
  Sample_ID = c("L1.1","L1.2","L1.3","L1.4","L1.5","L1.6","L1.7","L1.8","L1.9",
                "L3.10","L3.11","L3.12","L3.13","L3.14","L3.15","L3.16","L3.17","L3.18",
                "L5.19","L5.20","L5.21","L5.22","L5.23","L5.24","L5.25","L5.26","L5.27"),
  Location = c(rep("Location 1", 9), rep("Location 3", 9), rep("Location 5", 9)),
  Treatment = c(rep("CL", 3), rep("DD", 3), rep("MC", 3),
                rep("CL", 3), rep("DD", 3), rep("MC", 3),
                rep("CL", 3), rep("DD", 3), rep("MC", 3))
)

#====================================
#Phyloseq object
#====================================
otu_mat <- as.data.frame(raw_data[, 9:ncol(raw_data)])
rownames(otu_mat) <- raw_data[[1]] 
OTU <- otu_table(as.matrix(otu_mat), taxa_are_rows = TRUE)

tax_mat <- as.data.frame(raw_data[, 1:8])
rownames(tax_mat) <- tax_mat[[1]]
tax_mat_final <- as.matrix(tax_mat[, -1])
TAX <- tax_table(tax_mat_final)

metadata_df <- as.data.frame(metadata)
rownames(metadata_df) <- metadata_df$Sample_ID
META <- sample_data(metadata_df)


print(colnames(OTU))
print(rownames(META))
setdiff(colnames(OTU), rownames(META))

ps <- phyloseq(OTU, TAX, META)
saveRDS(ps, "ps_raw.rds")


```

# Filtering & transformation
```{r}
#Prevalence filtering
core_data <- core(x = ps,
                  detection = 0.01,
                  prevalence = 10/100)
core_data


```

# ALPHA DIVERSITY
Location-Treatment wise shannon diversity and observed richness is plotted

```{r}
#Alpha diversity
rich <- estimate_richness(ps, measures = c("Observed","Shannon"))
rich
write.csv(rich, "alpha_div/alpha_div.csv")

comparisons <- list(
  c("CL","DD"),
  c("CL","MC"),
  c("DD","MC")
)

p_sha <- plot_richness(ps, x = "Treatment", measures = "Shannon") +
  geom_boxplot(aes(fill = Treatment), outlier.shape = NA) + 
  facet_wrap(~Location) + 
  scale_x_discrete(labels = NULL) +
  theme_minimal()  +
  labs(y = "Shannon entropy", x = NULL) +
  theme(legend.position = "bottom",
    axis.text.x = element_blank(),
        axis.text = element_text(size = 12),
        axis.title.y = element_text(size = 12)
        ) + theme(
          strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    
    panel.spacing = unit(0, "lines"),
    
    panel.border = element_rect(color = "black", fill = NA, linetype = "dotted"),
    axis.text.x = element_text(angle = 45, hjust = 1, 
                                   size = 12),
        axis.text = element_text(size = 12),
        axis.title.y = element_text(size = 12)
        ) +
  scale_fill_manual(values = prism_cb_safe)


p_sha<- p_sha + stat_compare_means(method = "wilcox.test", 
                       comparisons = comparisons,
                       label = "p.format")

ggsave(p_sha, file = "alpha_div/shannon_entropy.png",bg = "white", width = 8, height = 6, dpi = 300)


print(p_obs)
print(p_sha)

```
# Wilcox Test

```{r}
rich_df<- as.data.frame(rich)
rich_df$Location <- sample_data(ps)$Location
rich_df$Treatment <- sample_data(ps)$Treatment
```
## Location wise wilcox test
```{r}

kr_loc_obs<- kruskal.test(Observed ~ Location, data = rich_df)
kr_loc_shn<- kruskal.test(Shannon ~ Location, data = rich_df)

wilcox_loc_obs <- pairwise.wilcox.test(rich_df$Observed, rich_df$Location, p.adjust.method = "fdr")
print(wilcox_loc_obs)

wilcox_loc_shannon <- pairwise.wilcox.test(rich_df$Shannon, rich_df$Location, p.adjust.method = "fdr")
print(wilcox_loc_shannon)
```
## treatment wise wilcox test
```{r}
kr_treat_obs <- kruskal.test(Observed ~ Treatment, data = rich_df)
kr_treat_shn <- kruskal.test(Shannon ~ Treatment, data = rich_df)

wilcox_observed <- pairwise.wilcox.test(
  rich_df$Observed,
  rich_df$Treatment,
  p.adjust.method = "fdr"
)

wilcox_shannon <- pairwise.wilcox.test(
  rich_df$Shannon,
  rich_df$Treatment,
  p.adjust.method = "fdr")
  
```
## Saving files
```{r}
#Kruskallis
test_results <- list()

test_df <- tibble(
  Type = c("Location","Location","Treatment","Treatment"),
  Test = c("Observed","Shannon","Observed","Shannon"),
  Statistic = c(kr_loc_obs$statistic,kr_loc_shn$statistic,
                kr_treat_obs$statistic,kr_treat_shn$statistic),
  krus_df = c(kr_loc_obs$parameter,kr_loc_shn$parameter,
              kr_treat_obs$parameter,kr_treat_shn$parameter),
  krus_p_value = c(kr_loc_obs$p.value,kr_loc_shn$p.value,
                   kr_treat_obs$p.value,kr_treat_shn$p.value)
)

write.csv(test_df, "alpha_div/kruskal.csv", row.names = FALSE)

# Wilcox
df_loc_obs <- tidy(wilcox_loc_obs) %>%
  mutate(Metric = "Observed", Type = "Location")

df_loc_shn <- tidy(wilcox_loc_shannon) %>%
  mutate(Metric = "Shannon", Type = "Location")

df_treat_obs <- tidy(wilcox_observed) %>%
  mutate(Metric = "Observed", Type = "Treatment")

df_treat_shn <- tidy(wilcox_shannon) %>%
  mutate(Metric = "Shannon", Type = "Treatment")

df_wilcox <- bind_rows(df_loc_obs,
                       df_loc_shn,
                       df_treat_obs,
                       df_treat_shn)

write_csv(df_wilcox,"alpha_div/wilcox_results.csv")

```

## Data Transformation
```{r}
dir.create("beta_div")

if(taxa_are_rows(core_data)){
  otu_mat <- t(otu_mat)
}
otu_nozero <- zCompositions::cmultRepl(otu_mat)
otu_clr <- compositions::clr(otu_nozero)

OTU_CLR <- otu_table(t(otu_clr), taxa_are_rows = TRUE)

ps_clr <- phyloseq(OTU_CLR,
                   tax_table(core_data),
                   sample_data(core_data))

```
##CLR transformation
```{r}
dist_mat <- phyloseq::distance(ps_clr, method = "euclidean")


ord <- ordinate(ps_clr,
                method = "PCoA",
                distance = "euclidean")
```

#PERMANOVA

```{r}
metadata_df <- as(sample_data(ps_clr), "data.frame")

adonis_result_overall <- adonis2(dist_mat ~ Location * Treatment,
                                 data = metadata_df,
                                 permutations = 999,
                                 by = "margin")

print(adonis_result_overall)
```

## Location-wise
```{r}
locations <- unique(metadata_df$Location)
adonis_result <- list()

for (loc in locations){

  phy_loc <- subset_samples(ps_clr, Location == loc)
  phy_loc <- prune_taxa(taxa_sums(phy_loc) > 0, phy_loc)

  dist_loc <- phyloseq::distance(phy_loc, method = "euclidean")
  meta_loc <- as(sample_data(phy_loc), "data.frame")

  ad <- adonis2(dist_loc ~ Treatment,
                data = meta_loc,
                permutations = 999)

  df <- as.data.frame(ad)
  df$Term <- rownames(df)
  df$Location <- loc

  adonis_result[[loc]] <- df
  print(adonis_result[[loc]])
}


```
## Treatment-wise pairwise adonis test

```{r}
metadata_df$Treatment <- as.factor(metadata_df$Treatment)

treatments <- unique(metadata_df$Treatment)

pairs <- combn(treatments, 2, simplify = FALSE)

pairwise_results <- list()

for(p in pairs){

  meta_sub <- metadata_df[metadata_df$Treatment %in% p, ]

  dist_sub <- as.dist(as.matrix(dist_mat)[rownames(meta_sub),
                                          rownames(meta_sub)])

  ad <- adonis2(dist_sub ~ Treatment, data = meta_sub)

  df <- as.data.frame(ad)
  df$Term <- rownames(df)
  df$Comparison <- paste(p, collapse = "_vs_")

  pairwise_results[[paste(p, collapse = "_vs_")]] <- df
}

pairwise_adonis_df <- do.call(rbind, pairwise_results)

write.csv(pairwise_adonis_df,
          "beta_div/pairwise_treatment_adonis.csv",
          row.names = FALSE)
print(pairwise_adonis_df)
```
#BETA DIVERSITY - Aitchison 
```{r}
metadata_df$Group <- paste(metadata_df$Location,
                           metadata_df$Treatment,
                           sep = "_")

disp <- betadisper(dist_mat, metadata_df$Group)

disp_df <- data.frame(
  SampleID = names(disp$distances),
  Distance = disp$distances
)
disp_df <- cbind(metadata_df,
                 Distance = disp$distances)

write.csv(disp_df, "beta_div/aitchison_dispersion.csv")

bray_box <- ggplot(disp_df,
                   aes(x = Treatment,
                       y = Distance,
                       fill = Treatment)) +
  geom_boxplot(aes(fill = Treatment),
               outlier.shape = NA) + 
  facet_wrap(~Location) + 
  scale_x_discrete(labels = NULL) +
  theme_bw() + stat_compare_means(method = "wilcox.test", 
                       comparisons = comparisons,
                       label = "p.format") +
  labs(
    y = "Aitchison distance",
    x = NULL
  ) +
    theme(legend.position = "bottom",
    axis.text.x = element_blank(),
        axis.text = element_text(size = 12),
        axis.title.y = element_text(size = 12)
        ) + theme(
          strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    panel.spacing = unit(0, "lines"),
    panel.border = element_rect(color = "black", fill = NA, linetype = "dotted"),
    axis.text.x = element_text(angle = 45, hjust = 1, 
                                   size = 12),
        axis.text = element_text(size = 12),
        axis.title.y = element_text(size = 12)
        ) +
  scale_fill_manual(values = prism_cb_safe)

bray_box + stat_compare_means(method = "wilcox.test", 
                       comparisons = comparisons,
                       label = "p.format")

  
ggsave(bray_box, file = "beta_div2/aitchison.png",bg = "white", width = 8, height = 6, dpi = 300)
bray_box

```
##Dispersion test

```{r}
anova(disp)
```

